### 2018-03-30 paper/sexSelectionSpacing-ver1.tex
Incorporate discussion of contraceptives and spacing; also minor edits

At a broader level, we know less about what determines spacing behavior in 
developing countries than in developed countries.
With increasing numbers of women entering the labor force in developing
countries, understanding how couples make timing decisions will be necessary
for the design of suitable policies \citep{Portner2018}.
Furthermore, if spacing does affect health outcomes for mother and children, it is 
essential to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arises
from sex-selective abortions, rather than, as often assumed, be driven by a 
higher value placed on girls because of their lower number \citep{Lin2014,Hu2015}.

@@

To overcome this problem, I condition on the likelihood of parity progression 
when examining birth spacing.
If, for example, women with a given set of characteristics have an 80\%
chance of ending a spell with a birth their median spell duration is when
their survival curve shows that 40\% are predicted to have exited the spell.%
\footnote{
It is important to note that this is not the same as simply calculating 
the median \emph{observed} spell lengths among women who have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
Median durations can then be averaged across women with different 
characteristics using the probabilities of ending a spell with a child
as weights.
How well this approach works depends on whether the spell length covered is 
sufficiently long that few women would be likely to give birth later.
I discuss the choice of spell length below.
In addition to calculating conditional spell length for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

@@

I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
As an example, for the third spell, three groups are used: Two boys,
one girl and one boy, and two girls.
[TK this should be more specific - more boys less sex selection - maybe even use ordering]
As discussed, the sex composition of previous children affects both the timing
of births and the use of sex-selective abortions.

@@

First, did the spacing after girls increase relative to the spacing
after boys?
The underlying hypothesis is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
available.
I still expect parents to \emph{try} to conceive earlier after the birth 
of a girl than after a boy, but the abortions will increase the 
length of the space, and may even lead to longer spacing after the
birth of girls than after boys.
[TK something on general changes in spacing? Or hold off until discussing the
results]

Second, do the shapes of survival curves differ across sex compositions 
within a spell?
The shapes of the survival curves provide clues to whether parents are
still trying to conceive earlier after girls than boys.
I expect this to be the case if there is still a strong preference for
boys, and, as discussed above, this behavior may even become more pronounced
with access to sex selection.
At a technical level, the shapes of the survival curve serve as an
indicator for the presence of non-proportionality.

@@

The spacing between births increases substantially over time, especially
for urban women. 
Urban women show an increase in median duration of between a quarter
and three-quarters of a year, whereas the increase for rural women is 
around a quarter of a year.
The spacing between marriage and the first birth is still the
exception, and for this group of women, there is a decline of
about a month for both urban and rural areas.

@@

The duration from marriage to first birth is only affected by son preferences if
people use sex selection to achieve a son as their first-born.
As I discuss above, the biased sex ratio for the first spell for the most educated 
women does suggest that there might be sex selection on the first birth.
I, therefore, begin by comparing in Figure \ref{fig:pss_spell1} how the distribution 
of spacing has changed over time for rural women with no education and urban women 
with eight or more years of education.
What is most striking is how similar the distribution of spacing is across the
two groups.

@@

Over the last 40 years, there has been a general increase in median
spacing for all education groups, but especially for the most 
educated women.
The exception is for the spacing between marriage and first birth
where there are two counteracting forces: women have become more likely 
to conceive after marriage, and there has been a substantial compression 
in the variation.
The result is that median spacing from marriage to first birth
that has changed little or even declined for the best-educated women.

@@

The results here also lead to a set of broader questions that
future research should address.
First, to what extent are the improvements in health for girls
relative to boys the result of selection, the longer spacing between
births, or changing son preference?
There is some early evidence that selection makes it appear that
girls are healthier, even though the real cause is a combination
of sex selection and higher mortality together with recall error \citep{Portner2018a}. 
Understanding what role these factors play in better health for
girls is especially important when evaluating policies that 
aim to directly limit the use of sex selection rather than 
changing preferences or incentives.

Second, we need to understand how female labor force participation
interacts with the use of sex selection.
Increased autonomy for women, arising from, for example, better opportunities
for working outside the home, has been suggested as a way to increase
women's status and thereby lower the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is an clear benefit to women who gain  
bargaining power, and it also increases the cost of repeated 
sex-selective abortions.
On the other hand, it may further lower desired fertility, and 
everything else equal, that may lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in
women's labor force participation and short-term costs from potential 
increases in sex selection is of paramount importance.

[Nothing is ever simple!]

[from introduction]
Furthermore, if spacing does affect health outcomes for mother and children, it is 
essential to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arises
from sex-selective abortions, rather than, as often assumed, be driven by a 
higher value placed on girls because of their lower number \citep{Lin2014,Hu2015}.

[Stuff to add somewhere in the paper]

In the absence of sex selection, I expect most parities to follow
a pattern, where there initially are relatively few births, 
followed by a substantial number of births over a 1 to 2 year period, and then
relatively few births thereafter.
As sex selection becomes more widely used the associated
longer spacing shows up by making the survival curve straighter,
indicating that some of the births that would originally have taken place
now take place later because of abortions.
The use of sex selection is clearly not the only factor that can change the shape of
the survival curves; factors such as the desired number of children and
use of contraceptives may also shape the shape.
To account for this it is best to compare survival curves for an individual
parity between women who are likely to use sex selection and women who are
not, for example because they already have achieved their desired 
number of sons.

[TK not sure where to fit this into]
Finally, increased reliability of access and effectiveness of
contraceptive can lead to shorter spacing between births 
\citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level 
of contraception---resulting in longer spacing---to avoid having 
too many children by accident.
But, as contraception becomes more effective parents can more
easily avoid future births, allowing them to reduce the spacing 
between births without having to worry about overshooting their 
preferred number of children.
This idea may also help explain shorter spacing for better 
educated women than for less educated women, provided that   
knowledge and ability to use contraception differ
across education groups \citep{Tulasidhar1993,Whitworth2002}.
It may also explain why finding statistically significant
effects of contraception use on birth spacing is difficult \citep{Yeakey2009}.

>>>>

At a broader level, we know less about what determines spacing behavior in 
developing countries than in developed countries.%
\footnote{
Even the effects of access to contraceptive on spacing are uncertain,
On one hand, access to contraceptive allow women to avoid too 
short spacing between birth, which would increase birth spacing.
On the other hand, increased reliability of access and effectiveness of
contraceptive can lead to shorter spacing between births 
\citep{Keyfitz1971,Heckman1976}.
With less reliable contraception parents choose a higher level 
of contraception---resulting in longer spacing---to avoid having 
too many children by accident.
But, as contraception becomes more effective parents can more
easily avoid future births, allowing them to reduce the spacing 
between births without having to worry about overshooting their 
preferred number of children.
This idea may explain why better educated women tha
shorter spacing than less educated women in certain 
cases \citep{Tulasidhar1993,Whitworth2002}.
These two counteracting effects may explain why finding statistically 
significant effects of contraception use on birth spacing is 
difficult \citep{Yeakey2009}.
}
With increasing numbers of women entering the labor force in developing
countries, understanding how couples make timing decisions will be necessary
for the design of suitable policies \citep{Portner2018}.
Furthermore, if spacing does affect health outcomes for mother and children, it is 
essential to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arises
from sex-selective abortions, rather than, as often assumed, be driven by a 
higher value placed on girls because of their lower number \citep{Lin2014,Hu2015}.

@@

To overcome this problem, I condition on the likelihood of parity progression 
when examining birth spacing.
If, for example, women with a given set of characteristics have an 80\%
chance of ending a spell with a birth their median spell duration is when
their survival curve shows that 40\% are predicted to have exited the spell.
It is important to note that this is not the same as simply calculating 
the median \emph{observed} spell lengths among women who have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.

Median durations can then be averaged across women using the probabilities 
of ending a spell with a child as weights.
How well this approach works depends on whether the spell length covered is 
sufficiently long that few women are be likely to give birth after
the spell cut-off.
I discuss the choice of spell length below.
In addition to calculating conditional spell length for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

@@

I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
As an example, for the third spell, three groups are used: Two boys,
one girl and one boy, and two girls.
As discussed, the sex composition of previous children affects both the timing
of births and the use of sex-selective abortions.

@@

First, did the spacing after girls increase relative to the spacing
after boys?
The underlying hypothesis is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
available.
I still expect parents to \emph{try} to conceive earlier after the birth 
of a girl than after a boy, but the abortions will increase the 
length of the space, and may even lead to longer spacing after the
birth of girls than after boys.

Second, have the shapes of the survival curves changed?
The shapes of the survival curves provide clues both to general
changes in birth spacing and to whether parents are
still trying to conceive earlier after girls than boys.
I expect this to be the case if there is still a strong preference for
boys, and, as discussed above, this behavior may even become more pronounced
with access to sex selection.
At a technical level, the shapes of the survival curve serve as an
indicator for the presence of non-proportionality.

@@

Birth spacing increases substantially over time, 
especially for urban women. 
The median duration for urban women increase between a quarter
and three-quarters of a year, whereas the increase for rural women 
is around a quarter of a year.
The spacing between marriage and the first birth is still the
exception, and for this group of women, there is a decline of
about a month for both urban and rural areas.

@@

The duration from marriage to first birth is only affected by son preferences if
people use sex selection to achieve a son as their first-born.
As I discuss above, the biased sex ratio for the first spell for the most educated 
women does suggest that there might be sex selection on the first birth.
I, therefore, begin by comparing in Figure \ref{fig:pps_spell1} how the distribution 
of spacing has changed over time for rural women with no education and urban women 
with eight or more years of education.
What is most striking is how similar the distribution of spacing is across the
two groups.

@@

Over the last 40 years, there has been a general increase in median
spacing for all education groups, but especially for the most 
educated women.
The exception is for the spacing between marriage and first birth
where there are two counteracting forces: women have become less likely 
to conceive before marriage, and there has been a substantial compression 
in the variation of spacing.
The result is that median spacing from marriage to first birth
that has changed little or even declined for the best-educated women.
Both the decrease for the first spell and the increase for higher-order
spells are consistent with an increased access to contraceptives over
time \citep{Yeakey2009}.

@@

The results here also lead to a set of broader questions that
future research should address.
First, to what extent are the improvements in health for girls
relative to boys the result of selection, the longer spacing between
births, or changing son preference?
There is some early evidence that sample selection can make it appear that
girls are healthier, even though the underlying cause is a combination
of sex selection and higher mortality together with recall 
error \citep{Portner2018a}. 
Understanding what role these factors play in better health for
girls is especially important when evaluating policies that 
aim to directly limit the use of sex selection rather than 
changing preferences or incentives.
If the better health outcomes for girls are, for example, an unintended 
side-effect of the longer spacing that arises from sex-selective 
abortions then an effective ban on sex selection may, at least
temporarily, worsen health outcomes for girls.

Second, we need to understand how female labor force participation
interacts with the use of sex selection.
Increased autonomy for women, for example from better opportunities
for working outside the home, has been suggested as a way to increase
women's status and thereby lower the use of sex selection \citep{Das-Gupta2016}.
This may, however, be a double-edged sword.
On the one hand, it is an clear benefit to the women who gain  
bargaining power, and it also increases the cost of repeated 
sex-selective abortions.
On the other hand, it may further lower desired fertility, and 
everything else equal, that may lead to higher use of sex selection.
Understanding the trade-off between long-term benefits from improvements in
women's labor force participation and short-term costs from potential 
increases in sex selection is of paramount importance.
